### 2018-03-01 contents/paper1.tex
Updates from ShareLaTeX

\graphicspath{{thesis/images/mspaper/drafts/}}

@@

\section{Introduction}
Trigeminal neuralgia is a neuropathic pain disorder characterized by severe, lancinating, facial pain with no major clinical sensory deficits. The pain occurs in one or more territories of the branches of the 5th cranial nerve (CN V) and is often triggered by innocuous stimuli, such as a light touch to the face. Classic trigeminal neuralgia is presumed to occur as a result of neurovascular compression at the root entry zone (REZ) of CN V (also referred as ‘idiopathic’; henceforth referred to simply as TN)(M Hodaie \& Coello, 2013). TN pain symptoms can also occur secondary to multiple sclerosis (MS-TN), where MS patients have a 20-fold risk of developing TN pain than the general population (up to 0.1\%)\cite{VanHecke2014}. 
While experientially similar, the pathophysiology of MS-TN differs from TN and involves CNS demyelination\cite{lazar1979trigeminal,Nurmikko2009}. The manner in which CNS demyelination results in pain is unclear. Brainstem plaques can be common in MS, but not all MS patient with brainstem plaques suffer from MS-TN. Furthermore, the presence of bilateral plaques does not clearly correlate with bilateral MS-TN pain\cite{DeSanti2011a,Love2001a}, highlighting the complex and uncertain relationship between the presence of plaques and pain\cite{DaSilva2005}. Importantly, this also highlights the fact that conventional MR plaque visualization does not fully correlate with MS clinical symptoms or their severity, and so improved neuroimaging methods are needed\cite{Seixas2014}. As classic TN is most commonly associated with vascular compression at the trigeminal nerve root entry zone, it is possible that in some MS patients neurovascular compression can also be present, adding further complexity to the diagnosis and treatment of trigeminal pain\cite{Sandell2010e}.

Understanding the microstructural anatomy of the trigeminal fibres is essential to determine how MS affects CN V brainstem fibres and how alterations to CN V in MS are related to the clinical presentation of pain. Direct comparisons between TN and MS-TN at the level of CN V brainstem fibres has been challenging because it is difficult to reliably identify, segment, and measure trigeminal anatomy with conventional magnetic resonance imaging (MRI)(Figure 1A-B)\cite{Cruccu2009,DaSilva2005}. Conventional MR imaging cannot visualize in isolation the brainstem fibres of CN V. Furthermore conventional tractography cannot resolve the brainstem crossing fibres\cite{Farquharson2013}, which leads to unreliable CN V delineations within the brainstem, and inaccuracies in diffusivity measurements.

Diffusion tensor imaging (DTI) metrics, based on estimating the Gaussian diffusion of water in the neural tissue, can provide important insights to microstructural neural tissue changes in both pathological and normal states. Metrics including fractional anisotropy (FA), and radial (RD), axial (AD), and mean diffusivities (MD), reflect diffusivity change measured as a parameter of the estimated Gaussian diffusion tensor eigenvectors.  FA reflects the shape of the ellipsoid Gaussian tensor as a measure of anisotropy, and is correlated with both structural and cognitive changes that follow adverse brain changes such as traumatic brain injury, aging and Alzheimer’s disease\cite{Bendlin2008,Charlton2006,Kinnunen2011c,Mielke2012}. RD is a measure of the cross-sectional diffusion as an average of the second and third diffusion eigenvalues, and has been shown to positively correlate with axonal demyelination \cite{Song2002,Song2005}. AD is the greatest diffusion eigenvalue and has been positively correlated with axonal degeneration\cite{Budde2009,Song2002}, and MD is the mean of all three diffusion eigenvalues and is related to the increase in tissue inflammation and edema\cite{Beaulieu2002}. In MS studies, FA, RD and MD appear abnormal in or near MS lesions. The changes in diffusivity alter the Gaussian tensor profile and results in decreased FA, increased RD and increased MD\cite{Braley2012,Janve2013,Klawiter2011,Lin2007,Senda2012g}. While AD is generally not correlated with demyelination\cite{Song2002}, there have been findings associating decreases in AD with the presence of neurological damage in MS\cite{Budde2008,Kim2006}.

The major technical limitation of diffusion tractography to date has been the inability of resolving crossing fibres using a single tensor Gaussian diffusion model\cite{Farquharson2013}, which has proved unsuitable for the visualization of brainstem tracts. However, techniques that use high angular resolution imaging  with an increased number of gradient directions acquired during diffusion weighted imaging (DWI), and advances in multi-tensor tractography (MTT) have improved the ability to delineate fibres in brain regions that have dense fibre crossings, such as the brainstem \cite{DellAcqua2007,Descoteaux2009c,Fillard2011,Qazi2009}. The microstructure of CN V in TN has recently been shown to be abnormal in terms of DTI metrics in the trigeminal REZ\cite{Desouza2013}. DTI metrics, as correlates of microstructural anatomy, are also important in identifying the nerve in pain in TN\cite{Hodaie2009a}.

@@

Images were analyzed with 3D Slicer (NA-MIC©, http://www.slicer.org)\cite{Pieper2004}. DWI sequences were corrected for eddy-current and motion distortions with appropriate corrections to gradient vectors, and imported into 3D Slicer for single-tensor diffusion tractography (SDT), where tracts were generated with initial seed spacing=0.5 mm, initial seeding FA threshold=0.2, stopping FA threshold=0.1, curvature threshold=0.8 and integration distance=0.1. Region-of-interests (ROI) seeding points were generated at a spacing of 0.25mm, or 64 points per voxel; MTT propagations were initiated at C1 threshold=0.2, tensor fraction=0.2, curve radius=0.8 rad, minimal length=10 mm, step size=1 mm. T$_{1}$ anatomical images were registered to the DWI using rigid transformation. To generate CN V tractography, ROIs were placed bilateral at the CN V at the REZ, and were used for both SDT and MTT.

@@

Statistical analysis were performed using R statistics software suite(R Development Core Team, 2012). Segments and diffusivity measurements were analyzed separately. Patient group is considered an independent and between-subject factor; symptomatic and asymptomatic sides within the same patient are considered within-subject measures; in controls, the mean measures of left/right sides are used for comparison against other groups. A linear mixed model\cite{Pinheiro2011} is used to determine the statistical significance of within- and between-subject effects. Post hoc analyses were performed using pair-wise Student’s t-test, where multiple comparisons are corrected with false discovery rate correction\cite{Hothorn2008}. Statistical results were plotted\cite{Hunter2007} as diffusivity difference estimate matrices between anatomical regions and different subject groups. Values are mapped as red/blue based on mixed linear regression estimates. Positive estimates, where the group diffusivity parameter is greater in the vertical axis versus the horizontal axis are in red, and negative coefficients are in blue.

@@

The results have agreed with our initial hypothesis. The coefficient matrices (Figures 4-7) revealed specific patterns demonstrating TN-specific diffusivity changes in the cistern and MS-TN-specific changes in the peri-lesional regions. In the TN REZ, FA changes are similar to the results found by DeSouza et al (Desouza, Moayedi, Chen, Davis, \& Hodaie, 2013; DeSouza et al., 2014), where the FA of the TN(+; pain-affected side) CN V is lower than that of TN(-; non-affected side). Additionally, while RD and MD changes in the TN REZ showed significant differences, no AD differences in the REZ were detected, suggesting that the nature of the diffusivity changes may be due to demyelination of the central myelin within the REZ. Interestingly, the cisternal TN(+) FA is demonstrably higher than the TN(-) segment, and also the cisternal FA of all other groups, suggesting that cisternal and REZ FA changes in TN are not homogenous along the nerve segment. Although there has been reports of hyperintensities in the cisternal CN V T$_{1}$ image with gadolinium contrast in MS \cite{VanderMeijs2002}, the lack of MS cistern findings in this study may be due to the limit of the sampled MS population. MS plaque locations are highly variable, and MS-TN patients were not filtered specifically for cistern region plaques.

Diffusivity disruptions of CN V in MS-TN patients are likely due to MS plaques at the regions proximal to the main sensory nucleus. While AD has been associated with neurological disruptions in MS\cite{Budde2009,Kim2006}, the present study showed a lack of change in AD. This may relate to the observation that we did not select subjects based on their disease severity, but based on their pain. Disease severity was in fact variable in this group. Future studies will be able to address this point in more detail, by taking into account other MS manifestations in addition to pain, and correlation with diffusion metrics. Lower REZ FA in MS-TN compared to controls suggests that global neural tissue diffusivity in MS-TN is significantly different from that of healthy controls. Similar findings can be seen in the pontine region, where only the MS-TN AD measures are distinctly different from the other groups. However while AD is commonly correlated with axonal integrity, the high concentration of crossing fibers in the pontine region may introduce confounds into these correlations. Improvements in diffusivity metric measurements in such tissues are needed to further understand these findings.

>>>>

\graphicspath{{images/mspaper/}}

@@

\section{Introduction}
Trigeminal neuralgia is a neuropathic pain disorder characterized by severe, lancinating, facial pain with no major clinical sensory deficits. The pain occurs in one or more territories of the branches of the 5th cranial nerve (CN V) and is often triggered by innocuous stimuli, such as a light touch to the face. Classic trigeminal neuralgia is presumed to occur as a result of neurovascular compression at the root entry zone (REZ) of CN V (also referred as ‘idiopathic’; henceforth referred to simply as TN)(M Hodaie \& Coello, 2013). TN pain symptoms can also occur secondary to multiple sclerosis (MS-TN), where MS patients have a 20-fold risk of developing TN pain than the general population (up to 0.1\%) \cite{VanHecke2014}. 
While experientially similar, the pathophysiology of MS-TN differs from TN and involves CNS demyelination \cite{lazar1979trigeminal,Nurmikko2009}. The manner in which CNS demyelination results in pain is unclear. Brainstem plaques can be common in MS, but not all MS patient with brainstem plaques suffer from MS-TN. Furthermore, the presence of bilateral plaques does not clearly correlate with bilateral MS-TN pain \cite{DeSanti2011a,Love2001a}, highlighting the complex and uncertain relationship between the presence of plaques and pain\cite{DaSilva2005}. Importantly, this also highlights the fact that conventional MR plaque visualization does not fully correlate with MS clinical symptoms or their severity, and so improved neuroimaging methods are needed \cite{Seixas2014}. As classic TN is most commonly associated with vascular compression at the trigeminal nerve root entry zone, it is possible that in some MS patients neurovascular compression can also be present, adding further complexity to the diagnosis and treatment of trigeminal pain \cite{Sandell2010e}.

Understanding the microstructural anatomy of the trigeminal fibres is essential to determine how MS affects CN V brainstem fibres and how alterations to CN V in MS are related to the clinical presentation of pain. Direct comparisons between TN and MS-TN at the level of CN V brainstem fibres has been challenging because it is difficult to reliably identify, segment, and measure trigeminal anatomy with conventional magnetic resonance imaging (MRI)(Figure 1A-B) \cite{Cruccu2009,DaSilva2005}. Conventional MR imaging cannot visualize in isolation the brainstem fibres of CN V. Furthermore conventional tractography cannot resolve the brainstem crossing fibres \cite{Farquharson2013}, which leads to unreliable CN V delineations within the brainstem, and inaccuracies in diffusivity measurements.

Diffusion tensor imaging (DTI) metrics, based on estimating the Gaussian diffusion of water in the neural tissue, can provide important insights to microstructural neural tissue changes in both pathological and normal states. Metrics including fractional anisotropy (FA), and radial (RD), axial (AD), and mean diffusivities (MD), reflect diffusivity change measured as a parameter of the estimated Gaussian diffusion tensor eigenvectors.  FA reflects the shape of the ellipsoid Gaussian tensor as a measure of anisotropy, and is correlated with both structural and cognitive changes that follow adverse brain changes such as traumatic brain injury, aging and Alzheimer’s disease \cite{Bendlin2008,Charlton2006,Kinnunen2011c,Mielke2012}. RD is a measure of the cross-sectional diffusion as an average of the second and third diffusion eigenvalues, and has been shown to positively correlate with axonal demyelination \cite{Song2002,Song2005}. AD is the greatest diffusion eigenvalue and has been positively correlated with axonal degeneration \cite{Budde2009,Song2002}, and MD is the mean of all three diffusion eigenvalues and is related to the increase in tissue inflammation and edema \cite{Beaulieu2002}. In MS studies, FA, RD and MD appear abnormal in or near MS lesions. The changes in diffusivity alter the Gaussian tensor profile and results in decreased FA, increased RD and increased MD \cite{Braley2012,Janve2013,Klawiter2011,Lin2007,Senda2012g}. While AD is generally not correlated with demyelination \cite{Song2002}, there have been findings associating decreases in AD with the presence of neurological damage in MS \cite{Budde2008,Kim2006}.

The major technical limitation of diffusion tractography to date has been the inability of resolving crossing fibres using a single tensor Gaussian diffusion model \cite{Farquharson2013}, which has proved unsuitable for the visualization of brainstem tracts. However, techniques that use high angular resolution imaging  with an increased number of gradient directions acquired during diffusion weighted imaging (DWI), and advances in multi-tensor tractography (MTT) have improved the ability to delineate fibres in brain regions that have dense fibre crossings, such as the brainstem \cite{DellAcqua2007,Descoteaux2009c,Fillard2011,Qazi2009}. The microstructure of CN V in TN has recently been shown to be abnormal in terms of DTI metrics in the trigeminal REZ \cite{Desouza2013}. DTI metrics, as correlates of microstructural anatomy, are also important in identifying the nerve in pain in TN\cite{Hodaie2009a}.

@@

Images were analyzed with 3D Slicer (NA-MIC©, http://www.slicer.org) \cite{Pieper2004}. DWI sequences were corrected for eddy-current and motion distortions with appropriate corrections to gradient vectors, and imported into 3D Slicer for single-tensor diffusion tractography (SDT), where tracts were generated with initial seed spacing=0.5 mm, initial seeding FA threshold=0.2, stopping FA threshold=0.1, curvature threshold=0.8 and integration distance=0.1. Region-of-interests (ROI) seeding points were generated at a spacing of 0.25mm, or 64 points per voxel; MTT propagations were initiated at C1 threshold=0.2, tensor fraction=0.2, curve radius=0.8 rad, minimal length=10 mm, step size=1 mm. T$_{1}$ anatomical images were registered to the DWI using rigid transformation. To generate CN V tractography, ROIs were placed bilateral at the CN V at the REZ, and were used for both SDT and MTT.

@@

Statistical analysis were performed using R statistics software suite(R Development Core Team, 2012). Segments and diffusivity measurements were analyzed separately. Patient group is considered an independent and between-subject factor; symptomatic and asymptomatic sides within the same patient are considered within-subject measures; in controls, the mean measures of left/right sides are used for comparison against other groups. A linear mixed model \cite{Pinheiro2011} is used to determine the statistical significance of within- and between-subject effects. Post hoc analyses were performed using pair-wise Student’s t-test, where multiple comparisons are corrected with false discovery rate correction \cite{Hothorn2008}. Statistical results were plotted \cite{Hunter2007} as diffusivity difference estimate matrices between anatomical regions and different subject groups. Values are mapped as red/blue based on mixed linear regression estimates. Positive estimates, where the group diffusivity parameter is greater in the vertical axis versus the horizontal axis are in red, and negative coefficients are in blue.

@@

The results have agreed with our initial hypothesis. The coefficient matrices (Figures 4-7) revealed specific patterns demonstrating TN-specific diffusivity changes in the cistern and MS-TN-specific changes in the peri-lesional regions. In the TN REZ, FA changes are similar to the results found by DeSouza et al \cite{Desouza2013,Desouza2013c}, where the FA of the TN(+; pain-affected side) CN V is lower than that of TN(-; non-affected side). Additionally, while RD and MD changes in the TN REZ showed significant differences, no AD differences in the REZ were detected, suggesting that the nature of the diffusivity changes may be due to demyelination of the central myelin within the REZ. Interestingly, the cisternal TN(+) FA is demonstrably higher than the TN(-) segment, and also the cisternal FA of all other groups, suggesting that cisternal and REZ FA changes in TN are not homogenous along the nerve segment. Although there has been reports of hyperintensities in the cisternal CN V T$_{1}$ image with gadolinium contrast in MS \cite{VanderMeijs2002}, the lack of MS cistern findings in this study may be due to the limit of the sampled MS population. MS plaque locations are highly variable, and MS-TN patients were not filtered specifically for cistern region plaques.

Diffusivity disruptions of CN V in MS-TN patients are likely due to MS plaques at the regions proximal to the main sensory nucleus. While AD has been associated with neurological disruptions in MS \cite{Budde2009,Kim2006}, the present study showed a lack of change in AD. This may relate to the observation that we did not select subjects based on their disease severity, but based on their pain. Disease severity was in fact variable in this group. Future studies will be able to address this point in more detail, by taking into account other MS manifestations in addition to pain, and correlation with diffusion metrics. Lower REZ FA in MS-TN compared to controls suggests that global neural tissue diffusivity in MS-TN is significantly different from that of healthy controls. Similar findings can be seen in the pontine region, where only the MS-TN AD measures are distinctly different from the other groups. However while AD is commonly correlated with axonal integrity, the high concentration of crossing fibers in the pontine region may introduce confounds into these correlations. Improvements in diffusivity metric measurements in such tissues are needed to further understand these findings.
